---
title: "Mouse Development Eye Creation"
output:
 html_notebook:
  author: "David McGaughey"
  date: "`r Sys.Date()`"
  theme: flatly
  toc: true
  toc_float: true
  code_folding: show
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,  warning = FALSE,
  collapse = TRUE,
  fig.width = 12, fig.height = 8,
  comment = "#>",
  dpi=300
)
```
# Stage 1
Pull in mouse cell metadata and retain eye cells.

Output the mouse ocular barcodes.

```{r, exec = FALSE}
library(tidyverse)
sample_meta <- data.table::fread('~/git/scEiaD_quant/sample_meta.scEiaD_v1.2025_02_03.02.tsv.gz')
cell_meta <- data.table::fread('~/data/scEiaD_modeling/mm111.adata.solo.20250131.obs.csv.gz')[,-1] %>% 
  relocate(barcode) %>% 
  filter(solo_doublet == "FALSE")

mm111_eye <- cell_meta %>% 
  filter(
    organ == 'Eye',
    organism == 'Mus musculus',
    !grepl("^#", sample_accession),
    source == 'Tissue')

#mm111_eye$barcode %>% write(gzfile('~/git/scEiaD_modeling/data/mm111_eye_bcs.20250221.csv.gz'))
```

```{b2 ct-prediction, eval = FALSE}
sinteractive --mem=128G --time=8:00:00  --gres=gpu:a100:1
cd /data/OGVFB_BG/scEiaD/2024_02_28/snakeout/mm111_developing_eye
source /data/$USER/conda/etc/profile.d/conda.sh && source /data/$USER/conda/etc/profile.d/mamba.sh
mamba activate rapids_singlecell
bash ct_projection_call.sh
```



## scanvi Modeling
### scEiaD Human Developing Model
```{r}
sceiad_meta <- fst::read_fst('~/data/scEiaD_2022_02/meta_filter.fst')
ctp_mm111__human_modeling <- data.table::fread('/Users/mcgaugheyd/data/scEiaD_modeling/mm111_developing_eye/mm111.eye.human_CT_projections_20120128.csv.gz') %>% select(-17) %>% left_join(sceiad_meta %>% select(barcode = Barcode, CellType_predict), by = 'barcode') %>% 
  # only look at >= 10 day cells
  mutate(age = as.numeric(age)) %>% 
  filter(age < 10)

label <- 'MajorCellType'
machine_label <- 'CT__sceiad_20250211_dev_mmGeneFilter'
most_common_mislabel <- ctp_mm111__human_modeling %>% 
  group_by(.data[[label]],.data[[machine_label]]) %>% 
  summarise(Count = n()) %>% 
  arrange(.data[[label]], -Count) %>% 
  slice_max(order_by = Count, n = 2) %>% 
  filter(.data[[label]] != .data[[machine_label]])

ctp_mm111__human_modeling %>% 
  mutate(tf = case_when(.data[[label]] == .data[[machine_label]] ~ TRUE,
                        TRUE ~ FALSE)) %>% 
  group_by(.data[[label]]) %>% 
  summarise(accuracy = sum(tf)/length(tf)) %>% 
  arrange(.data[[label]]) %>% 
  left_join(most_common_mislabel %>% 
              select(any_of(label), 
                     `most common mislabel` = any_of(machine_label),
                     `mislabel count` = Count))
```




#### UMAP
```{r, fig.width=12,fig.height=12}
umap1 = 'umap1_sceiad_20250211_dev_mmGeneFilter'
umap2 = 'umap2_sceiad_20250211_dev_mmGeneFilter'
ct = 'CT__sceiad_20250211_dev_mmGeneFilter'
color = 'CT__sceiad_20250211_dev_mmGeneFilter'
score <- 'CT__sceiad_20250211_dev_mmGeneFilter__max_score'
umap_ct_plotter <- function(obj, umap1, umap2, ct, color, labels = TRUE){
  plot <- obj %>% 
    mutate(leiden = as.factor(leiden)) %>% 
    ggplot(aes(x=.data[[umap1]],y=.data[[umap2]])) +
    scattermore::geom_scattermore(aes(color = .data[[color]]), pointsize = 0.8, alpha = 0.5) +
    scale_color_manual(values = c(pals::alphabet2(), pals::glasbey(), pals::kelly(), pals::alphabet()) %>%
                         unname()) + 
    cowplot::theme_cowplot() + theme(legend.position = "none")
  if (labels){
    plot <- plot + ggrepel::geom_label_repel(data = . %>% 
                                               group_by(.data[[ct]]) %>% 
                                               summarise(across(all_of(c(umap1,umap2)), median)),
                                             aes(label = .data[[ct]], color = .data[[color]])) 
    
  } 
  print(plot)
}

umap_score_plotter <- function(obj, umap1, umap2, ct, score){
  obj %>% 
    ggplot(aes(x=.data[[umap1]],y=.data[[umap2]])) +
    scattermore::geom_scattermore(aes(color = .data[[score]]), pointsize = 0.8, alpha = 0.5) +
    ggrepel::geom_label_repel(data = . %>% 
                                group_by(.data[[ct]]) %>% 
                                summarise(across(all_of(c(umap1,umap2)), median)),
                              aes(label = .data[[ct]]), max.overlaps = Inf) + 
    scale_color_viridis_c() +
    cowplot::theme_cowplot() + theme(legend.position = "none")
}
umap_ct_plotter(ctp_mm111__human_modeling, umap1, umap2, ct, color)
umap_ct_plotter(ctp_mm111__human_modeling, umap1, umap2, 'leiden', 'leiden')
umap_score_plotter(ctp_mm111__human_modeling, umap1, umap2, ct, score)
```


## MajorCellType <-> CT__sceiad_20250211_dev_mmGeneFilter
Looks quite good - perhaps only pericyte and ciliary margin (neither of which are in the big model) are wrong?
```{r}
ctp_mm111__human_modeling %>% 
  mcHelpeRs::sum_rat('CellType_predict', 'MajorCellType', 'CT__sceiad_20250211_dev_mmGeneFilter', threshold = 0.05) %>% 
  #filter(MajorCellType != '') %>% 
  DT::datatable()
```
## Rough set CT calls for scVI modeling
As a crucial step is subsetting of each CT to prevent a very high n set (e.g. rods) to dominate
```{r}
ctp_mm111__human_modeling <- ctp_mm111__human_modeling %>% 
  mutate(rough_ct = case_when(CellType_predict %in% c("Ciliary Margin", "Pericyte") ~ tolower(CellType_predict),
                              TRUE ~ CT__sceiad_20250211_dev_mmGeneFilter))
```

# Stage 2
Pick ref barcodes to make scVI models

```{r}
set.seed(20250303)
keep_ct <- ctp_mm111__human_modeling %>% group_by(rough_ct) %>% summarise(Count = n()) %>% filter(Count > 10) %>% pull(rough_ct) %>% unique() %>% sort()


mm111_dev_eye_ref_bcs <- ctp_mm111__human_modeling %>% ungroup() %>% 
  filter(rough_ct %in% keep_ct) %>% 
  group_by(study_accession, rough_ct) %>% 
  sample_n(1000, replace = TRUE) %>% 
  unique()

mm111_dev_eye_query_bcs <- ctp_mm111__human_modeling %>% 
  ungroup() %>% 
  filter(!barcode %in% mm111_dev_eye_ref_bcs$barcode)

#mm111_dev_eye_ref_bcs$barcode %>% write(gzfile('~/git/scEiaD_modeling/data/mm111_dev_eye_ref_bcs.20250303.csv.gz'))
#mm111_dev_eye_query_bcs$barcode %>% write(gzfile('~/git/scEiaD_modeling/data/mm111_dev_eye_query_bcs.20250303.csv.gz'))
```

```{bash b2 scvi time, eval = FALSE}
cd /data/OGVFB_BG/scEiaD/2024_02_28/snakeout/mm111_developing_eye
sbatch --time=10:00:00 snakecall.sh
```

```{r}
source('analysis_scripts.R')
obs <- pull_obs('~/data/scEiaD_modeling/mm111_developing_eye/mm111_dev_eye_20250303_2000hvg_50e_30l.obs.csv.gz', machine_label = 'scANVI_MCT', cluster_col = 'leiden5')

obs$obs <- obs$obs %>% left_join(ctp_mm111__human_modeling %>% select(barcode, CT__sceiad_20250211_dev_mmGeneFilter, CT__sceiad_20250211_dev_mmGeneFilter__max_score,  umap1_sceiad_20250211_dev_mmGeneFilter,umap2_sceiad_20250211_dev_mmGeneFilter), by = c("barcodei" = 'barcode'))

obs$labels$mCT_scvi <- obs$obs %>% mcHelpeRs::sum_rat("leiden5","CT__sceiad_20250211_dev_mmGeneFilter") %>% slice_max(order_by = Count, n = 1) %>% pull(CT__sceiad_20250211_dev_mmGeneFilter)

obs$labels$mMCT_scvi <- obs$obs %>% mcHelpeRs::sum_rat("leiden5","CT__sceiad_20250211_dev_mmGeneFilter") %>% summarise(out = paste0(CT__sceiad_20250211_dev_mmGeneFilter, collapse = ', ')) %>% pull(out)
```
## UMAPs
```{r, fig.width=12, fig.height=12}
obs$obs %>% 
  left_join(obs$labels, by = 'leiden5') %>% 
  ggplot(aes(x=umap1,y=umap2)) +
  scattermore::geom_scattermore(aes(color = scANVI_MCT), pointsize = 0.8, alpha = 0.5) +
  ggrepel::geom_label_repel(data = . %>% group_by(scANVI_MCT) %>% 
                              summarise(umap1 = median(umap1),
                                        umap2 = median(umap2)),
                            aes(label = scANVI_MCT, color = scANVI_MCT)) +
  scale_color_manual(values = c(pals::alphabet2(), pals::glasbey()) %>% unname()) + 
  cowplot::theme_cowplot() + theme(legend.position = "none") +
  ggtitle("MajorCellType (scANVI)")

obs$obs %>% 
  left_join(obs$labels, by = 'leiden5') %>% 
  ggplot(aes(x=umap1,y=umap2)) +
  scattermore::geom_scattermore(aes(color = CT__sceiad_20250211_dev_mmGeneFilter), pointsize = 0.8, alpha = 0.5) +
  ggrepel::geom_label_repel(data = . %>% group_by(CT__sceiad_20250211_dev_mmGeneFilter) %>% 
                              summarise(umap1 = median(umap1),
                                        umap2 = median(umap2)),
                            aes(label = CT__sceiad_20250211_dev_mmGeneFilter, color = CT__sceiad_20250211_dev_mmGeneFilter)) +
  scale_color_manual(values = c(pals::alphabet2(), pals::glasbey()) %>% unname()) + 
  cowplot::theme_cowplot() + theme(legend.position = "none") +
  ggtitle("CT__sceiad_20250211_dev_mmGeneFilter")  

obs$obs %>% 
  left_join(obs$labels, by = 'leiden5') %>% 
  ggplot(aes(x=umap1,y=umap2)) +
  scattermore::geom_scattermore(aes(color = CT__sceiad_20250211_dev_mmGeneFilter), pointsize = 0.8, alpha = 0.5) +
  scale_color_manual(values = c(pals::alphabet2(), pals::glasbey()) %>% unname()) + 
  cowplot::theme_cowplot() + theme(legend.position = "none") +
  ggtitle("CT__sceiad_20250211_dev_mmGeneFilter") +
  coord_cartesian(xlim = c(-20,20), ylim = c(-15,15)) +
  facet_wrap(~study_accession)

obs$obs %>% 
  left_join(obs$labels, by = 'leiden5') %>% 
  mutate(age = as.integer(age)) %>% 
  ggplot(aes(x=umap1,y=umap2)) +
  scattermore::geom_scattermore(aes(color = age), pointsize = 0.8, alpha = 0.5) +
  # ggrepel::geom_label_repel(data = . %>% group_by(CT__sceiad_20250211_dev_mmGeneFilter) %>% 
  #                             summarise(umap1 = median(umap1),
  #                                       umap2 = median(umap2)),
  #                           aes(label = CT__sceiad_20250211_dev_mmGeneFilter)) +
  scale_color_viridis_c(option = 'H') + 
  cowplot::theme_cowplot() +
  ggtitle("CT__sceiad_20250211_dev_mmGeneFilter") +
  coord_cartesian(xlim = c(-20,20), ylim = c(-15,15))
```

## leiden5 UMAP
```{r, fig.width=12, fig.height=12}
obs$obs %>%
  left_join(obs$labels, by = 'leiden5') %>% 
  ggplot(aes(x=umap1,y=umap2)) +
  scattermore::geom_scattermore(aes(color = as.factor(leiden5)), pointsize = 0.8, alpha = 0.5) +
  ggrepel::geom_text_repel(data = . %>% group_by(mCT_scvi, leiden5) %>%
                             summarise(umap1 = median(umap1),
                                       umap2 = median(umap2),),
                           aes(label = paste0(mCT_scvi,'-',leiden5),color = as.factor(leiden5)), bg.color = 'white') +
  scale_color_manual(values = c(pals::alphabet2(), pals::glasbey(), pals::alphabet(), pals::kelly()) %>% unname()) + 
  cowplot::theme_cowplot() + theme(legend.position = "none") +
  ggtitle("leiden5") 
```


## hclust
```{r, fig.width = 18, fig.height = 10}
pb <- data.table::fread('/Users/mcgaugheyd/data/scEiaD_modeling/mm111_developing_eye/mm111_dev_eye_20250303_2000hvg_50e_30l.pseudoBulk.leiden5.csv.gz')
colnames(pb) <- gsub("\\.\\d+","",colnames(pb))
hvg <- data.table::fread('/Users/mcgaugheyd/data/scEiaD_modeling/mm111_developing_eye/hvg2000.csv.gz')[-1,]
rnames <- pb$V1
clust <- str_extract(rnames, '\\d+') %>% as.integer()
pb <- pb[,-1] %>% as.matrix()
row.names(pb) <- as.character(clust)
pb <- pb[as.character(obs$labels$leiden5),]

pb_norm <- metamoRph::normalize_data(t(pb), sample_scale = 'cpm') %>% t() 

pb_full <- pb_norm
pb_norm <- pb_norm[,hvg$V2]

#pb_norm <- pb_norm[,hvg$V2[!hvg$V2 %in% c(cc_genes,ribo_genes)]]
# https://stats.stackexchange.com/questions/31565/compute-a-cosine-dissimilarity-matrix-in-r
sim <- pb_norm / sqrt(rowSums(pb_norm * pb_norm))
sim <- sim %*% t(sim)
D_sim <- as.dist(1 - sim)

hclust_sim <- hclust(D_sim, method = 'average')

hclust_sim$labels <- obs$labels %>% pull(leiden5)

library(ggtree)
p <- ggtree(hclust_sim)

p$data <- p$data %>% left_join(obs$label, by = c("label" = "leiden5"))
p + layout_dendrogram() +
  geom_tiplab(aes(label = paste(label, mCT_scvi, sep = ' - '), color = mCT_scvi)) + 
  theme_dendrogram(plot.margin=margin(16,16,300,16)) +
  scale_color_manual(values = c(pals::alphabet2(), pals::glasbey()) %>% unname()) + 
  guides(color="none")
```

# CT by CT
```{r}
diff <- pull_diff("~/data/scEiaD_modeling/mm111_developing_eye/mm111_dev_eye_20250303_2000hvg_50e_30l.difftesting.leiden5.csv.gz", orgdb =  org.Mm.eg.db::org.Mm.eg.db, cluster_col = 'leiden5')

conv_table <- AnnotationDbi::select(org.Mm.eg.db::org.Mm.eg.db,
                                    keys=gsub('\\.\\d+','',unique(diff$diff_testing$ENSEMBL)),
                                    columns=c("ENSEMBL","SYMBOL", "GENENAME", "ENTREZID"), keytype="ENSEMBL")

```



```{r, fig.width=20, fig.height=5}
library(ComplexHeatmap)

hm_maker <- function(markers, target, 
                     cdiff = diff, 
                     clabels = labels, 
                     remove = remove_leiden5){
  tib <- cdiff$diff_testing %>% 
    left_join(clabels, by = c('base'='leiden5')) %>% 
    left_join(conv_table %>% select(SYMBOL, ENSEMBL) %>% unique()) %>% 
    filter(SYMBOL %in% markers) %>% 
    mutate(base = as.character(base),
           base = paste0(base, ' - ', CT)) %>% 
    select(SYMBOL, base, logfoldchanges) %>% 
    pivot_wider(values_from = logfoldchanges, names_from = base)
  
  mat <- tib %>% select(-1) %>% as.matrix()
  row.names(mat) <- tib %>% pull(1)
  
  ha_column = ComplexHeatmap::HeatmapAnnotation(df = data.frame(Target = ifelse(grepl(target, colnames(tib)[-1]), "Target","Not"),
                                                                Remove = ifelse(str_extract(colnames(tib)[-1], '\\d+') %in% remove, "Remove","Retain")),        
                                                col = list(Target = c("Target" = "black","Not" = "white"),
                                                           Remove = c("Remove" = "red", "Retain" = "white")))
  
  col_fun = circlize::colorRamp2(c(-3, 0, 3), c("blue", "white", "red"))
  draw(Heatmap(mat, col=col_fun,
               name = 'logFoldChange',
               top_annotation = ha_column)
  )
}
```
## Quick Check of Discrepancies when filtering out lower confidence scores
```{r}
mct = obs$obs %>% group_by(leiden5, CT__sceiad_20250211_dev_mmGeneFilter) %>% count() %>% ungroup() %>% group_by(leiden5) %>% slice_max(order_by = n, n = 1)
filter_mct = obs$obs %>% filter(CT__sceiad_20250211_dev_mmGeneFilter__max_score > 0.9) %>% group_by(leiden5, CT__sceiad_20250211_dev_mmGeneFilter) %>% count() %>% ungroup() %>% group_by(leiden5) %>% slice_max(order_by = n, n = 1)
mct %>% left_join(filter_mct, by = 'leiden5') %>% filter(CT__sceiad_20250211_dev_mmGeneFilter.x != CT__sceiad_20250211_dev_mmGeneFilter.y)
```
## CT Changes
```{r}
ct_map <- c(
  '21' = 'cone (ml)',
  '61' = 'periocular mesenchyme',
  '62' = 'ciliary margin',
  '65' = 'rod (precursor)',
  '85' = 'lens',
  '77' = 'astrocyte',
  '13' = 'retinal ganglion',
  '44' = 'retinal ganglion',
  '51' = 'retinal ganglion',
  '47' = 'horizontal',
  '14' = 'neurogenic', # check above with score discrepancy 
  '87' = 'neurogenic', # check above with score discrepancy
  '70' = 'bipolar (precursor)' ) # check above with score discrepancy

labels <- obs$labels %>% 
  mutate(CT = ifelse(as.character(leiden5) %in% names(ct_map), ct_map[as.character(leiden5)], mCT_scvi))

remove_leiden5 <- c(43,74) # odd umap and marker expression (mixed ct)
```

## Heatmaps
### RPC / Neurogenic
```{r, fig.width=20, fig.height=7}
markers <- c("HES1",
             "ZFP36L2",
             # "HES6",
             # "ATOH7",
             "VIM",
             "CCND1",
             "SFRP2",
             "SPP1",
             "ZFP36L1",
             "TF",
             "FOS",
             "TTYH1") %>% str_to_title()
mellough_markers <- read_csv("~/git/eyeMarkers/lists/rpc_markers__Mellough2019.csv")

remove_leiden5 <- c()

markers <- mellough_markers %>% filter(`Cell Type` == 'RPC') %>% pull(HGNC) %>% str_to_title()
more <- c("PAX6","NEUROD1","ATOH7","HES6") %>% str_to_title()
hm_maker(c(markers, more), "rpc|neuro", clabels = obs$labels %>% mutate(CT = mCT))

```




### Periocular Mesenchyme / Endo / Epi / Keratocyte
```{r, fig.width=20, fig.height=6}
markers <- c("LUM","DCN","VIM","PDGFRA","COL1A2", # https://www.nature.com/articles/s42003-020-0922-4
             "MGP","MEG3","DCN","APOD","ANGPTL7","EFEMP1","BMP5","PRRX1")

markers <- #c("VIM","FAP","COL1A1","PDGFRB","S100A4", # fibro
  c("PENK", "PITX2", #POM
    "CDH5","VWF", # endo
    "MYF5", "USP18", #connective tissue
    "CDH1","KRT19","EPCAM", # epi
    "KERA",# keratocyte)
    "A2M") 
hm_maker(markers %>% str_to_title(), "periocu|endo|epi|kera",clabels = labels)
```

## Lens
```{r, fig.width=20, fig.height=12}
markers <- conv_table %>% filter(grepl("^Cry", SYMBOL)) %>% pull(SYMBOL)
hm_maker(markers %>% str_to_title(), "lens",clabels = labels)
```

## RPE
```{r, fig.width=20, fig.height=4}
markers <- c("Rpe65","Mitf","Best1")
hm_maker(markers %>% str_to_title(), "rpe",clabels = labels)
```

## Astrocyte
```{r, fig.width=20, fig.height=4}
markers <- c( "Gfap","Pax2")
hm_maker(markers %>% str_to_title(), "astroc",clabels = labels)
```

## Mueller
```{r, fig.width=20, fig.height=6}
markers <- c( "Glul","Clu","Apoe")
hm_maker(markers %>% str_to_title(), "astroc",clabels = labels)
```

## Horizontal
```{r, fig.width=20, fig.height=5}

markers <- c("LHX1","ONECUT1") %>% str_to_title()

hm_maker(markers, "hori")

```

## Photoreceptors
```{r, fig.width=20, fig.height=5}

markers <-  c('ARR3','OPN1LW','OPN1SW','RHO', 'OPN1MW', 'RCVRN',"CRX","PROM1","CNGA1","PDE6A") %>% str_to_title()
#markers <-  mellough_markers %>% filter(`Cell Type` %in% c('Rod','Cone')) %>% pull(HGNC)
hm_maker(markers, "rod|cone")

```


## Bipolar
```{r, fig.width=20, fig.height=5}
markers <- c("GRIK1","IRX6","LRTM1","PCP2","PRKCA","TRPM1","VSX1","VSX2") %>% str_to_title()
#markers <- mellough_markers %>% filter(`Cell Type` == 'Bipolar') %>% pull(HGNC)
hm_maker(markers, "bipolar")

```

## Amacrine
```{r, fig.width=20, fig.height=5}

markers <- c('GAD1','GAD2','SLC6A9','NFIA') %>% str_to_title()
markers <- mellough_markers %>% filter(`Cell Type` == 'Amacrine') %>% pull(HGNC) %>% str_to_title()
hm_maker(markers, "amacr")

```

## Ganglion

```{r, fig.width=20, fig.height=5}

markers <-  mellough_markers %>% filter(`Cell Type` == 'RGC') %>% pull(HGNC) %>% str_to_title()
hm_maker(markers, "ganglion")

```




# Updated UMAP

```{r, fig.width=12, fig.height=12}
nobs <- obs$obs %>% 
  left_join(labels, by = 'leiden5') %>% 
  filter(!leiden5 %in% remove_leiden5) 

# obs$obs %>% filter(leiden5 %in% c(14,57)) %>% mcHelpeRs::sum_rat(leiden5, CT__sceiad_20250211_dev_mmGeneFilter) 
# # tweaking for cluster 14 and 57 as these don't seem to be clearly bipolar precursors
# nobs <- nobs %>% mutate(CT = case_when(leiden5 == 14 & CT__sceiad_20250211_dev_mmGeneFilter == 'bipolar (precursor)' ~ 'rod (precursor)',
#                                        leiden5 == 14 & CT__sceiad_20250211_dev_mmGeneFilter == 'neurogenic' ~ 'neurogenic',
#                                        leiden5 == 14 & CT__sceiad_20250211_dev_mmGeneFilter == 'rod (precursor)' ~ 'rod (precursor)',
#                                        leiden5 == 14 & CT__sceiad_20250211_dev_mmGeneFilter == 'cone (precursor)' ~ 'rod (precursor)',
#                                        
#                                        leiden5 == 57 & CT__sceiad_20250211_dev_mmGeneFilter == 'bipolar (precursor)' ~ 'rod (precursor)',
#                                        leiden5 == 57 & CT__sceiad_20250211_dev_mmGeneFilter == 'rod (precursor)' ~ 'rod (precursor)',
#                                        leiden5 == 57 & CT__sceiad_20250211_dev_mmGeneFilter == 'neurogenic' ~ 'neurogenic',
#                                        leiden5 == 57 & CT__sceiad_20250211_dev_mmGeneFilter == 'bipolar' ~ 'bipolar',
#                                        leiden5 == 57 ~ 'neurogenic',
#                                        TRUE ~ CT
# ))
nobs %>% 
  ggplot(aes(x=umap1,y=umap2)) +
  scattermore::geom_scattermore(aes(color = CT), pointsize = 0.8, alpha = 0.5) +
  ggrepel::geom_label_repel(data = . %>% group_by(CT) %>% 
                              summarise(umap1 = median(umap1),
                                        umap2 = median(umap2)),
                            aes(label = CT, color = CT)) +
  scale_color_manual(values = c(pals::alphabet2(), pals::glasbey()) %>% unname()) + 
  cowplot::theme_cowplot() + theme(legend.position = "none") +
  ggtitle("CT")  
```

# Stage 3 
Output updated CT calls, update the h5ad, and re-run scVI

```{r}
set.seed(2025-03-05)
ref <- nobs %>% 
  filter(scANVI_MCT_max_score > 0.9) |> 
  group_by(study_accession, CT) %>% 
  slice_sample(n = 1000, replace = TRUE) %>% 
  unique()

query <- nobs %>% filter(!barcodei %in% ref$barcodei)


# ref$barcodei %>% write(gzfile('~/git/scEiaD_modeling/data/mm111_dev_eye_ref_bcs.20250305.stage3.csv.gz'))
# query$barcodei %>% write(gzfile('~/git/scEiaD_modeling/data/mm111_dev_eye_query_bcs.20250305.stage3.csv.gz'))
# #
# nobs %>% dplyr::rename(barcode = barcodei) %>% write_csv('~/git/scEiaD_modeling/data/Mouse_Developing_Eye__stage3_CTcalls.freeze20250305.01.csv.gz')
```

```{bash biowulf2 append obs, eval = FALSE}
cd /data/OGVFB_BG/scEiaD/2024_02_28/snakeout/hs111_developing_eye/stage4
source /data/$USER/conda/etc/profile.d/conda.sh && source /data/$USER/conda/etc/profile.d/mamba.sh
mamba activate rapids_singlecell

python ~/git/scEiaD_modeling/workflow/scripts/append_obs.py ../../mm111.adata.solo.20250131.h5ad /home/mcgaugheyd/git/scEiaD_modeling/data/Mouse_Developing_Eye__stage3_CTcalls.freeze20250305.01.csv.gz  mm111.adata.solo.20250305.dev.stage3.h5ad --transfer_columns CT
```

## obs s3
```{r}
obs_s3 <- pull_obs('~/data/scEiaD_modeling/mm111_developing_eye/stage3/mm111_dev_eye_20250305_stage3_2000hvg_200e_30l.obs.csv.gz', label = 'CT', machine_label = 'scANVI_CT', cluster_col = 'leiden5')
```

```{r, fig.width=9, fig.height=9}

obs_s3$obs %>%  ggplot(aes(x=umap1,y=umap2)) +
  scattermore::geom_scattermore(aes(color = CT), pointsize = 0.8, alpha = 0.5) +
  ggrepel::geom_label_repel(data = . %>% group_by(CT) %>% 
                              summarise(umap1 = median(umap1),
                                        umap2 = median(umap2)),
                            aes(label = CT, color = CT)) +
  scale_color_manual(values = c(pals::alphabet2(), pals::glasbey()) %>% unname()) + 
  cowplot::theme_cowplot() + theme(legend.position = "none") +
  ggtitle("CT")  



obs_s3$obs %>%
  left_join(obs_s3$labels, by = 'leiden5') %>% 
  ggplot(aes(x=umap1,y=umap2)) +
  scattermore::geom_scattermore(aes(color = as.factor(leiden5)), pointsize = 0.8, alpha = 0.5) +
  ggrepel::geom_text_repel(data = . %>% group_by(mCT, leiden5) %>%
                             summarise(umap1 = median(umap1),
                                       umap2 = median(umap2),),
                           aes(label = paste0(mCT,'-',leiden5),color = as.factor(leiden5)), bg.color = 'white') +
  scale_color_manual(values = c(pals::alphabet2(), pals::glasbey(), pals::alphabet(), pals::kelly(), pals::brewer.set1(10), pals::okabe(), pals::watlington()) %>% unname()) + 
  cowplot::theme_cowplot() + theme(legend.position = "none") +
  ggtitle("leiden5") 

```

## Confusion Matrix


```{r, fig.width=12, fig.height=12}
machine_label = 'scANVI_CT'; label = 'CT'
obs_s3$obs %>% 
  #filter(scANVI_CT_max_score > 0.9) %>% 
  group_by(.data[[label]],.data[[machine_label]]) %>% 
  summarise(Count = n()) %>% 
  mutate(Ratio = Count/sum(Count)) %>% 
  ggplot(aes(x=.data[[label]],y=.data[[machine_label]],fill=Ratio, label = round(Ratio, 2))) + 
  geom_tile() + 
  shadowtext::geom_shadowtext() + 
  cowplot::theme_cowplot() +
  scale_fill_viridis_c(begin = 0, end = 1) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

```{r}
# retains CT calls >= 0.05 of a cluster. anything below gets changed to the dominant ct
nobs_s3_cleaning <- obs_s3$obs %>% 
  group_by(leiden5, scANVI_CT) %>% 
  summarise(Count = n()) %>%
  mutate(Ratio = Count / sum(Count)) %>% 
  mutate(dominant_celltype = scANVI_CT[which.max(Count)]) %>% 
  mutate(CTc = case_when(Ratio < 0.05 ~ dominant_celltype,
                         TRUE ~ scANVI_CT))

nobs_s3 <- obs_s3$obs %>% 
  left_join(nobs_s3_cleaning %>% 
              dplyr::select(leiden5, scANVI_CT, CTc), by = c("leiden5","scANVI_CT"))
nobs_s3 <- nobs_s3 %>% select(-umap1, -umap2) %>% left_join(obs$obs %>% dplyr::select(umap1,umap2, barcodei), by = 'barcodei')

nobs_s3 %>% mcHelpeRs::sum_rat(CT, CTc,threshold = 0.01)
```

# Final UMAP
```{r, fig.width=9, fig.height=9}

nobs_s3 %>%  ggplot(aes(x=umap1,y=umap2)) +
  scattermore::geom_scattermore(aes(color = CTc), pointsize = 0.8, alpha = 0.5) +
  ggrepel::geom_label_repel(data = . %>% group_by(CTc) %>% 
                              summarise(umap1 = median(umap1),
                                        umap2 = median(umap2)),
                            aes(label = CTc, color = CTc)) +
  scale_color_manual(values = c(pals::alphabet2(), pals::glasbey()) %>% unname()) + 
  cowplot::theme_cowplot() + theme(legend.position = "none") +
  ggtitle("CTc")  
```
## "Transitional" 
Plot of the ML cell type call confidence (1 is 100% confidence). We see that in several "precursor" positions in the UMAP the confidence is much lower. 
```{r, fig.width=9, fig.height=9}
nobs_s3 %>%  ggplot(aes(x=umap1,y=umap2)) +
  scattermore::geom_scattermore(aes(color = scANVI_CT_max_score), pointsize = 0.8, alpha = 0.5) +
  ggrepel::geom_label_repel(data = . %>% group_by(CTc) %>% 
                              summarise(umap1 = median(umap1),
                                        umap2 = median(umap2)),
                            aes(label = CTc), color = 'black') +
  scale_color_viridis_c(option = 'rocket') + 
  cowplot::theme_cowplot() + 
  ggtitle("CTc")  
```


# Output
```{r}
#nobs_s3 %>%  
#  write_csv("~/data/scEiaD_modeling/mm111_developing_eye/stage3/mm111_dev_eye_20250305_stage3_2000hvg_200e_30l.nobs_s3.csv.gz")
```

```{r}
sessionInfo()
```