---
title: "Mouse Mature Backeye Creation"
output:
 html_notebook:
  author: "David McGaughey"
  date: "`r Sys.Date()`"
  theme: flatly
  toc: true
  toc_float: true
  code_folding: hide
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,  warning = FALSE,
  collapse = TRUE,
  fig.width = 12, fig.height = 8,
  comment = "#>",
  dpi=300
)
```

# Stage 1

Evaluate the cell type labels (the "truth" for scANVI / scVI). 

```{bash, exec = FALSE}
# edit and run this locally (laptop / etc)
~/git/scEiaD_quant/workflow/scripts/hand_change_ct_labels.R
# rsync the new cell label file to biowulf
rsync -Prav scEiaD_cell_labels_2024_08_27.csv.gz h2:/home/mcgaugheyd/git/scEiaD_quant/
# create the organism level h5ad file from the individual scEiaD_quant made h5ad files
# now on biowulf
sinteractive --mem=128G
cd /data/OGVFB_BG/scEiaD/2024_02_28/
mamba deactivate; mamba activate rscvi
python ~/git/scEiaD_modeling/workflow/scripts/merge_adata.py mm111.fin.txt /home/mcgaugheyd/git/scEiaD_quant/sample_meta.scEiaD_v1.2024_02_28.01.tsv.gz /home/mcgaugheyd/git/scEiaD_quant/scEiaD_cell_labels_2024_08_27.csv.gz mm111.adata.solo.20240827.h5ad mm111.adata.solo.20240827.obs.csv.gz
# back to local computer to copy the cell metadata
cd ~/git/scEiaD_modeling
rsync -Prav h2:/data/OGVFB_BG/scEiaD/2024_02_28/mm111.adata.solo.20240827.obs.csv.gz .
```

# Stage 2

- pull data from one species
- filter to age group (dev or mature)
- select random (up to 2k) cell type per study and output those barcodes
- output barcodes for the non-selected cells
- run scvi on biowulf

```{r, exec = FALSE}
library(tidyverse)
sample_meta <- data.table::fread('~/git/scEiaD_quant/sample_meta.scEiaD_v1.2024_02_28.01.tsv.gz')
cell_meta <- data.table::fread('~/data/scEiaD_modeling/mm111.adata.solo.20240827.obs.csv.gz')[,-1] %>% 
  relocate(barcode) %>% 
  filter(solo_doublet == "FALSE")

mm111_backeye <- cell_meta %>% 
  mutate(stage = case_when(as.numeric(age) <= 10 ~ 'Developing', 
                           TRUE ~ 'Mature'), 
         side = case_when(tissue == 'Brain Choroid Plexus' ~ 'Brain Choroid Plexus',
                          grepl("Choroid|RPE", tissue) ~ 'BackEye',
                          grepl("Retina", tissue) ~ 'BackEye',
                          grepl("Outf", tissue) ~ 'FrontEye',
                          grepl("Iris", tissue) ~ 'FrontEye',
                          grepl("Sclera", tissue) ~ 'FrontEye',
                          grepl("Cornea", tissue) ~ 'FrontEye',
                          grepl("Macula", tissue) ~ 'BackEye',
                          grepl("Trabecul", tissue) ~ 'FrontEye',
                          grepl("Optic", tissue) ~ 'BackEye',
                          TRUE ~ tissue)) %>% 
  filter(organism == 'Mus musculus',
         !grepl("^#", sample_accession),
         source == 'Tissue',
         #tissue %in% c("Macula", "Retina"),
         #side %in% c("FrontEye", "BackEye"),
         side %in% c("BackEye"),# 2024 08 26
         #capture_type == 'cell', # 2024 08 28
         #kb_tech %in% c("10xv1","10xv2","10xv3"), # 2024 08 28
         stage == 'Mature') %>% 
  filter(study_accession != 'SRP447468') # 2024 08 29 remove (for now????) this optic nerve and adjacent study from sanes as it clusters (hclust) apart from everything else

mm111_backeye$MajorCellType %>% table()
# do not use studies with less than 1000 cells as ref
mm111_study <- mm111_backeye %>% 
  group_by(study_accession) %>% summarise(Count = n()) %>% filter(Count>1000)
set.seed(101294)
mm111_ref_bcs <- mm111_backeye %>% 
  filter(study_accession %in% mm111_study$study_accession) %>% 
  filter(MajorCellType != '', MajorCellType != 'rpc') %>%  # 2024 04 16 big change - only use labelled cells in ref
                                    # 2024 07 30 fixed mistake in filtering that still let them in
  group_by(study_accession, MajorCellType) %>%  # 2024 07 30 big change - sample on study *and* majorcelltype
  sample_n(2000, replace = TRUE) %>%  
  unique() %>% 
  pull(barcode)

mm111_query_bcs <- mm111_backeye %>% 
  filter(!barcode %in% mm111_ref_bcs) %>% 
  pull(barcode)

mm111_ref_bcs %>% write(gzfile('~/git/scEiaD_modeling/data/mm111_mature_backeye_ref_bcs.20240830-02.csv.gz'))
mm111_query_bcs %>% write(gzfile('~/git/scEiaD_modeling/data/mm111_mature_backeye_query_bcs.20240830-02.csv.gz'))

```

## run scVI

now go to biowulf2:/data/OGVFB_BG/scEiaD/2024_02_28/snakeout/mm111_mature_backeye

```{bash, exec = FALSE}
# will take ~2-4 hours (depending on hpc usage)
mamba deactivate; mamba activate; bash ~/git/scEiaD_modeling/Snakemake.wrapper.sh ~/git/scEiaD_modeling/workflow/Snakefile ~/git/scEiaD_modeling/config/config_mm111_mature_backeye.yaml ~/git/scEiaD_modeling/config/cluster.json
```

## rsync output from biowulf2 to local computer
```{bash, exec = FALSE}
cd /Users/mcgaugheyd/data/scEiaD_modeling/mm111_mature_backeye
rsync -Prav h2:/data/OGVFB_BG/scEiaD/2024_02_28/snakeout/*backeye*/*scib* .
rsync -Prav h2:/data/OGVFB_BG/scEiaD/2024_02_28/snakeout/*backeye*/*obs* .
rsync -Prav h2:/data/OGVFB_BG/scEiaD/2024_02_28/snakeout/*backeye*/*.leiden3.csv.gz .
rsync -Prav h2:/data/OGVFB_BG/scEiaD/2024_02_28/snakeout/*backeye*/*hvg.csv.gz .
```

# Stage 3
## scib scoring
```{r}
scib_files <- list.files(path = '~/data/scEiaD_modeling/mm111_mature_backeye/', full.names = TRUE) %>% grep('scib',.,value = TRUE)
scib_scores <- purrr::map(scib_files, read.csv)
names(scib_scores) <- scib_files %>% str_extract("_\\d+hvg.*l") %>% gsub("^_","",.)
for (i in names(scib_scores)){
  scib_scores[[i]]$Delta <- as.numeric(scib_scores[[i]][2,'Total']) - as.numeric(scib_scores[[i]][1,'Total'])
}
bind_rows(scib_scores, .id = 'Params') %>% 
  filter(Embedding == 'X_scVI') %>% 
  mutate(Params= gsub("hvg|e|l","",Params)) %>% 
  separate(Params, into = c("HVG","Epochs", "Latent Dimensions"), sep = "_") %>% 
  mutate_all(as.numeric) %>% 
  mutate(Embedding = 'X_scVI') %>% 
  arrange(-Total)
```



## Assess Output
```{r}
obs <- data.table::fread('~/data/scEiaD_modeling/mm111_mature_backeye/mm111_mature_backeye_20240830_2000hvg_100e_50l.obs.csv.gz' )[,-1] %>% relocate(barcode) %>% 
  mutate(side = case_when(tissue == 'Brain Choroid Plexus' ~ 'Brain Choroid Plexus',
                          grepl("Choroid|RPE", tissue) ~ 'BackEye',
                          grepl("Retina", tissue) ~ 'BackEye',
                          grepl("Outf", tissue) ~ 'FrontEye',
                          grepl("Iris", tissue) ~ 'FrontEye',
                          grepl("Sclera", tissue) ~ 'FrontEye',
                          grepl("Cornea", tissue) ~ 'FrontEye',
                          grepl("Macula", tissue) ~ 'BackEye',
                          grepl("Trabecul", tissue) ~ 'FrontEye',
                          grepl("Optic", tissue) ~ 'BackEye',
                          TRUE ~ tissue)) 

obs_machine_labels <- obs %>% group_by(leiden3, MCT_scANVI) %>% 
  filter(MCT_scANVI != 'unlabelled') %>% 
  summarise(Count = n()) %>% mutate(Ratio = Count/sum(Count)) %>% 
  filter(Ratio > 0.10) %>% 
  arrange(-Ratio) %>% 
  summarise(mMCT = paste0(MCT_scANVI, collapse = ', '),
            mCount = sum(Count),
            mMCT_Ratio = paste0(round(Ratio,2), collapse =', '),
            mCT = gsub(", .*","",mMCT)) 

obs_author_labels <- obs %>% group_by(leiden3, MajorCellType) %>% 
  filter(MajorCellType != 'unlabelled') %>% 
  summarise(Count = n()) %>% mutate(Ratio = Count/sum(Count)) %>% 
  filter(Ratio > 0.10) %>% 
  arrange(-Ratio) %>% 
  summarise(aMCT = paste0(MajorCellType, collapse = ', '),
            aCount = sum(Count),
            aMCT_Ratio = paste0(round(Ratio,2), collapse =', '),
            aCT = gsub(", .*","",aMCT)) 

obs_leiden3_ct <- obs %>% group_by(leiden3) %>% 
  summarise(TotalCount = n(), 
            solo_score = mean(solo_score) %>% round(.,2), 
            background_fraction = mean(background_fraction) %>% round(.,2), 
            cell_probability = mean(cell_probability) %>% round(.,2))


obs_labels <- obs_author_labels %>% 
  left_join(obs_machine_labels, by = c("leiden3")) %>% 
  left_join(obs_leiden3_ct, by = 'leiden3') 

obs_study <- obs %>% group_by(leiden3, study_accession) %>% 
  summarise(studyCount = n()) %>% 
  filter(studyCount > 10) %>% 
  summarise(studyCount = n(), studies = paste0(study_accession, collapse = ', '))

obs_tissue <- obs %>% group_by(leiden3, tissue) %>% 
  summarise(tissueCount = n()) %>% 
  summarise(tissueCount = n(), tissues = paste0(unique(tissue), collapse = ', '))

obs_tech<- obs %>% group_by(leiden3, workflow) %>% 
  summarise(techCount = n()) %>% 
  mutate(techRatio = techCount/sum(techCount)) %>% 
  filter(workflow == 'standard') %>% 
  select(-techCount, -workflow) 
  

obs_labels <- obs_labels %>% 
  left_join(obs_study, by = 'leiden3') %>% 
  left_join(obs_tissue, by = 'leiden3') %>% 
  left_join(obs_tech, by = 'leiden3')
```

```{r diff}
# pull in diff
diff_testing <- data.table::fread("~/data/scEiaD_modeling/mm111_mature_backeye/mm111_mature_backeye_20240830_2000hvg_100e_50l.difftesting.leiden3.csv.gz") %>% mutate(ENSEMBL = gsub("\\.\\d+","",names))
conv_table <- AnnotationDbi::select(org.Mm.eg.db::org.Mm.eg.db,
                                    keys=gsub('\\.\\d+','',unique(diff_testing$names)),
                                    columns=c("ENSEMBL","SYMBOL","GENENAME", "ENTREZID"), keytype="ENSEMBL")

top_diff <- diff_testing %>% 
  group_by(base) %>% 
  slice_max(n = 7, order_by = scores) %>% 
  left_join(conv_table, by =c( "ENSEMBL")) %>% 
  ungroup() %>% 
  mutate(leiden3 = base) %>% 
  group_by(leiden3) %>% 
  summarise(diff_genes = paste0(SYMBOL, collapse = ', '))

obs_labels <- obs_labels %>% 
  left_join(top_diff)
```




## Cluster Overview

### Ratio (percentage) of labelled cell types for each leiden3 cluster
```{r}
obs_labels %>% 
  arrange(aCT) %>% 
  mutate(leiden3 = as.factor(leiden3)) %>% 
  DT::datatable(filter = 'top')
```



### background fraction against solo (doublet) score
Labelled points are clusters with more than one machine CT (above 0.1)
```{r}
obs_labels %>% 
  mutate(mMCT_unique_count = str_count(mMCT, ',') + 1) %>%  
  ggplot(aes(x=solo_score,y=background_fraction)) + 
  geom_point(aes(color = as.factor(mMCT_unique_count), size = log1p(TotalCount))) + 
  ggrepel::geom_label_repel(data = . %>% filter(mMCT_unique_count > 1), aes(label = leiden3), max.overlaps = Inf)
```


### Discrepancies
Between author cell type and machine (scANVI) cell type
```{r}
obs_labels %>% filter(aCT != mCT) %>% select(leiden3:TotalCount) %>% data.frame
```

Multiple mMCT above 10% of the total in a cluster
```{r}
obs_labels %>% filter(grepl(",",mMCT)) %>% select(leiden3:TotalCount) %>% data.frame %>% arrange(-TotalCount)
```

## UMAP Plots
```{r, fig.width=16, fig.height=16}
obs %>% 
  left_join(obs_labels, by = 'leiden3') %>% 
  ggplot(aes(x=umap1,y=umap2)) +
  scattermore::geom_scattermore(aes(color = mCT), pointsize = 0.8, alpha = 0.5) +
  ggrepel::geom_label_repel(data = . %>% group_by(mCT) %>% 
                              summarise(umap1 = median(umap1),
                                        umap2 = median(umap2)),
                            aes(label = mCT, color = mCT)) +
  scale_color_manual(values = c(pals::alphabet2(), pals::glasbey()) %>% unname()) + 
  cowplot::theme_cowplot() + theme(legend.position = "none")

obs %>% 
  left_join(obs_labels, by = 'leiden3') %>% 
  ggplot(aes(x=umap1,y=umap2)) +
  scattermore::geom_scattermore(aes(color = mMCT), pointsize = 4, alpha = 0.5) +
  scale_color_manual(values = c(pals::alphabet2(), pals::glasbey()) %>% unname()) + 
  cowplot::theme_cowplot() + theme(legend.position = "none") + 
  facet_wrap(~study_accession)



obs %>% 
  left_join(obs_labels, by = 'leiden3') %>% 
  ggplot(aes(x=umap1,y=umap2)) +
  scattermore::geom_scattermore(aes(color = as.factor(leiden3))) +
  facet_wrap(~mCT) +
  scale_color_manual(values = c(pals::alphabet2(), pals::glasbey(),pals::kelly(n=22),pals::alphabet(),pals::brewer.set1(n=10)) %>% unname()) + 
  cowplot::theme_cowplot()

obs %>% 
  left_join(obs_labels, by = 'leiden3') %>% filter(mMCT == 'bipolar') %>% 
  ggplot(aes(x=umap1,y=umap2)) +
  scattermore::geom_scattermore(aes(color = study_accession), pointsize = 0.8, alpha = 0.8) +
  ggrepel::geom_label_repel(data = . %>% group_by(mCT) %>% 
                              summarise(umap1 = median(umap1),
                                        umap2 = median(umap2)),
                            aes(label = mCT)) +
  scale_color_manual(values = c(pals::alphabet2(), pals::glasbey()) %>% unname()) + 
  cowplot::theme_cowplot() 

obs %>% 
  left_join(obs_labels, by = 'leiden3') %>% filter(mMCT == 'rpe') %>% 
  ggplot(aes(x=umap1,y=umap2)) +
  scattermore::geom_scattermore(aes(color = study_accession), pointsize = 0.8, alpha = 0.8) +
  ggrepel::geom_label_repel(data = . %>% group_by(mCT) %>% 
                              summarise(umap1 = median(umap1),
                                        umap2 = median(umap2)),
                            aes(label = mCT)) +
  scale_color_manual(values = c(pals::alphabet2(), pals::glasbey()) %>% unname()) + 
  cowplot::theme_cowplot() 

obs %>% 
  left_join(obs_labels, by = 'leiden3') %>% filter(mMCT == 'amacrine') %>% 
  ggplot(aes(x=umap1,y=umap2)) +
  scattermore::geom_scattermore(aes(color = study_accession), pointsize = 0.8, alpha = 0.8) +
  ggrepel::geom_label_repel(data = . %>% group_by(mCT) %>% 
                              summarise(umap1 = median(umap1),
                                        umap2 = median(umap2)),
                            aes(label = mCT)) +
  scale_color_manual(values = c(pals::alphabet2(), pals::glasbey()) %>% unname()) + 
  cowplot::theme_cowplot() 

obs %>% 
  left_join(obs_labels, by = 'leiden3') %>% filter(mMCT == 'amacrine') %>% 
  ggplot(aes(x=umap1,y=umap2)) +
  scattermore::geom_scattermore(aes(color = as.factor(leiden3)), pointsize = 0.8, alpha = 0.8) +
  ggrepel::geom_label_repel(data = . %>% group_by(mCT) %>% 
                              summarise(umap1 = median(umap1),
                                        umap2 = median(umap2)),
                            aes(label = mCT)) +
  scale_color_manual(values = c(pals::alphabet2(), pals::glasbey()) %>% unname()) + 
  cowplot::theme_cowplot() 
```

## hclust
Take pseudobulk values (at the cluster level) and hierarchically cluster them to ensure 
there aren't any issues in either the overall structure (e.g. rod and cones are intersperse)d
and/or to identify any potential mislabeled clusters

```{r, fig.width = 18, fig.height = 10}
library(tidyverse)
pb <- data.table::fread('~/data/scEiaD_modeling/mm111_mature_backeye/mm111_mature_backeye_20240830_2000hvg_100e_50l.pseudoBulk.leiden3.csv.gz')
hvg <- data.table::fread('~/data/scEiaD_modeling/mm111_mature_backeye/hvg500.csv.gz')[-1,]
rnames <- pb$V1
clust <- str_extract(rnames, '\\d+') %>% as.integer()
pb <- pb[,-1] %>% as.matrix()
row.names(pb) <- as.character(clust)
pb <- pb[as.character(obs_labels$leiden3),]

# library(SingleCellExperiment)
# library(scater)
# library(scran)
# do_hvg <- function(cts){
#   sce <- SingleCellExperiment(list(counts = cts))
#   sce <- logNormCounts(sce)
#   hvg <- modelGeneVar(sce)
#   # Visualizing the fit:
#   hvg$var <- metadata(hvg)$var
#   return(hvg)
# }
# 
# nhvg  <- do_hvg(t(pb))

pb_norm <- metamoRph::normalize_data(t(pb), sample_scale = 'cpm') %>% t() 

# remove cell cycle genes
conv_table <- AnnotationDbi::select(org.Mm.eg.db::org.Mm.eg.db, 
                                    keys=gsub('\\.\\d+','',unique(colnames(pb_norm))),
                                    columns=c("ENSEMBL","SYMBOL","GENENAME", "ENTREZID"), keytype="ENSEMBL")

cc_genes <- hvg %>% mutate(ENSEMBL = gsub("\\.\\d+","",V2)) %>% 
  left_join(conv_table, by = "ENSEMBL") %>% 
  mutate(cc_genes = case_when(SYMBOL %in% (Seurat::cc.genes.updated.2019 %>% unlist()) ~ TRUE)) %>% 
  filter(cc_genes) %>% pull(V2)
ribo_genes <- hvg %>% mutate(ENSEMBL = gsub("\\.\\d+","",V2)) %>% 
  left_join(conv_table, by = "ENSEMBL") %>% filter(grepl("^RPL|^RPS|^MT",SYMBOL)) %>% 
  pull(SYMBOL)

pb_norm <- pb_norm[,hvg$V2[!hvg$V2 %in% c(cc_genes,ribo_genes)]]
# https://stats.stackexchange.com/questions/31565/compute-a-cosine-dissimilarity-matrix-in-r
sim <- pb_norm / sqrt(rowSums(pb_norm * pb_norm))
sim <- sim %*% t(sim)
D_sim <- as.dist(1 - sim)

hclust_sim <- hclust(D_sim, method = 'average')

hclust_sim$labels <- obs_labels %>% pull(leiden3)

library(ggtree)
p <- ggtree(hclust_sim)
p$data <- p$data %>% left_join(obs_labels, by = c("label" = "leiden3")) %>% 
  mutate(techRatio = round(techRatio, digits = ))
p + layout_dendrogram() +
  geom_tiplab(aes(label = paste(label, mMCT, studyCount, TotalCount,sep = ' - '), color = mCT)) + 
  theme_dendrogram(plot.margin=margin(16,16,300,16)) +
  scale_color_manual(values = c(pals::alphabet2() %>% unname())) + 
  guides(color="none")
```
```{r, fig.width=10}
mct <- obs %>% group_by(MajorCellType, study_accession, tissue) %>% 
  filter(MajorCellType != '', MajorCellType != 'unlabelled', MajorCellType != 'rpc') %>% 
  summarize(Count = n()) %>% 
  rename(MajorCellType = 'CT') %>% 
  mutate(Class = 'Author Major Cell Type')

mMCT <- obs %>% group_by(MCT_scANVI, study_accession, tissue) %>% 
  filter(MCT_scANVI != '', MCT_scANVI != 'unlabelled') %>% 
  summarize(Count = n()) %>% 
  rename(MCT_scANVI = 'CT') %>% 
  mutate(Class = 'ML Major Cell Type')

bind_rows(mct,
          mMCT) %>% 
  ggplot(aes(y=CT,x=Count,fill=study_accession)) +
  geom_bar(stat='identity') +
  facet_grid(~Class) +
  cowplot::theme_cowplot() +
  scale_fill_manual(values = pals::kelly()[2:20])

```

```{r}
sessionInfo()
```